
In the context of patterns in medical images, computer vision models extract \textit{imaging features} that capture the underlying texture and shape of an image.  These imaging features have led to an explosion in machine learning tools that find common patterns within the features for inference, determining a cancerous lesion as benign or malignant, and generative tasks, synthesizing new medical images for augmenting data. , it is a natural step to exploit imaging features to find patterns across patient populations during diagnosis and progression.  Medical imaging has another benefit of

Furthermore, the necessity of repeated, invasive sampling and molecular assay may be burdensome to the patient, is expensive, and limits the practical number of opportunities to monitor disease progression and treatment response.   

A bottleneck of many of these imaging biomarkers is the reliance on the spatial relationship between pixels, where they are not powerful enough to distinguish between lesions for a prediction task or the modality does not generate pixels.  A workaround could utilize the power of neural networks as is the current direction of medical imaging, but limited datasets for some modalities are a bottleneck for inference tasks.   or the contrast within a lesion is too low for to generate imaging biomarkers that can discriminate for infeerence tasks.  We generate imaging biomarkers for and Positron Emission/Computed Tomography.    

%%The study of images is an inherent natural phenomena pervasive throughout history occupying discussions outside of the arts in fields such as math, medicine, and geography.  The common discourse in all of these fields is the expression of an image by it's patterns or as the sum of it's parts.  For example, in mathematics, classical Euclidean topology representations include a set of points in $\mathbb{R^{n}}$ and the accompanying Euclidean distance metrics to describe location and nearness.  Within medicine, the difference in contrast of a medical image helps segment tissue from bones.    



%%Cancer incidence and mortality rates have increased yearly and is still %%responsible for one in eight deaths worldwide %%[https://onlinelibrary.wiley.com/doi/full/10.3322/caac.21492].  


%%hh are omnipotent at the initial stage of patient diagnosis

%%diological tools such as standard uptake volume (SUV), the texture and shape image utilized in natural images can easily be defined for medical image and have been exploited in many applications ranging from computer-aided detection, diagnosis, and discrimination of disease.  Instead of imaging features they are often to referred to as imaging biomarkers.

%% Deleterious mutations in the EGFR gene [B. Rude Voldborg, EGFR] cause the inactivation of the extracelluar domain, rendering this domain constitutively active even in the absence of upstream events such as ligand-binding or dimerization.  